---
title: "Data Analysis and Visualization in R"
author: "Bibhash Nath"
date: "4/20/2021"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

## Set up a R project for the R-Spatial section. 

```{r}
library(sf)
library(sp)
library(rgdal)
require(ggmap)
library(tidyverse) #for tidyverse functions
library(janitor) # For cleaning up column names
```

## Read the NYC postal areas in Shapefiles into sf and sp objects. As NYC DOH publishes COVID-19 data by zip code, we will utilize the postal area data later.

```{r}
nyc_zc_sf <- st_read("ZIP_CODE_040114.shp")
class(nyc_zc_sf) # is a sf object
plot(nyc_zc_sf['ZIPCODE'])

nyc_zc_sp <- readOGR("ZIP_CODE_040114.shp")
class(nyc_zc_sp) # is a sp object
```

## Read and process the NYC public health services spreadsheet data. Create sf and sp objects from geographic coordinates.

```{r}
nys_hf <- read_csv("NYS_Health_Facility.csv") # reading health facility data
county <- c("Bronx", "Kings", "Queens", "New York", "Richmond") # subset for NYC
nyc_hf <- nys_hf %>% filter(`Facility County` %in% county) %>% 
  filter(`Facility Latitude` > 0) %>%
  drop_na(`Facility Latitude`) # subset for NYC, dropping NA and zero values

nyc_hf <- tidyr::separate(nyc_hf, `Facility Location`, into = c("Latitude", "Longitude"), sep = ",")
nyc_hf <- nyc_hf %>% mutate(Latitude = str_replace_all(Latitude, "\\*|\\(|\\)", ""))
nyc_hf <- nyc_hf %>% mutate(Longitude = str_replace_all(Longitude, "\\*|\\(|\\)", ""))
nyc_hf$Latitude <- as.numeric(as.character(nyc_hf$Latitude))
nyc_hf$Longitude <- as.numeric(as.character(nyc_hf$Longitude))
# above lines of code were used to clean latitude and longitude and make numeric.
```

```{r}
# sf objects for nyc health facility
nyc_hf_sf <- st_as_sf(nyc_hf, coords = c("Longitude", "Latitude"))
st_crs(nyc_hf_sf) <- 4326 # set CRS to WGS84
class(nyc_hf_sf) # is a sf object

# sp objects for nyc health facility
coordinates(nyc_hf) <- c("Longitude", "Latitude")
proj4string(nyc_hf) <- CRS("+init=epsg:4326")
proj4string(nyc_hf)
class(nyc_hf) # is a sp object
```

## Read and process the NYS retail food stores data. Create sf and sp objects from geographic coordinates for NYC.

```{r}
nys_rf <- read_csv("NYS_Retail_Food_Stores.csv")
nys_rf <- clean_names(nys_rf)
county_sub <- c("Bronx", "Kings", "Queens", "New York", "Richmond") # subset for NYC
nyc_rf <- nys_rf %>% filter(county %in% county_sub) # subset for NYC

nyc_rf <- tidyr::separate(nyc_rf, location, into = c("location", "clean"), sep = "[(]")
nyc_rf <- tidyr::separate(nyc_rf, clean, into = c("Latitude", "Longitude"), sep = ",")
nyc_rf <- nyc_rf %>% mutate(Longitude = str_replace_all(Longitude, "\\*|\\(|\\)", ""))
nyc_rf$Latitude <- as.numeric(as.character(nyc_rf$Latitude))
nyc_rf$Longitude <- as.numeric(as.character(nyc_rf$Longitude))
nyc_rf <- nyc_rf %>% filter(nyc_rf$Latitude > 40)
# above lines of code were used to clean latitude and longitude and make numeric.
```

```{r}
# sf objects for nyc retail stores
nyc_rf_sf <- st_as_sf(nyc_rf, coords = c("Longitude", "Latitude"))
st_crs(nyc_rf_sf) <- 4326 # set CRS to WGS84
class(nyc_rf_sf) # is a sf object

# sp objects for nyc retail stores
coordinates(nyc_rf) <- c("Longitude", "Latitude")
proj4string(nyc_rf) <- CRS("+init=epsg:4326")
proj4string(nyc_rf)
class(nyc_rf) # is a sp object
```

## Use simple mapping method with a basemap to verify the above datasets in terms of their geometry locations.

```{r}
# For NYC health facility
# verify sf objects
nyc_hf_sf %>% st_union() %>% st_bbox() %>% as.vector() %>%
  ggmap::get_stamenmap(zoom = 11) -> baseMap

ggmap(baseMap) +
  geom_point(aes(x=X, y=Y), 
             data = nyc_hf_sf %>% st_coordinates() %>% 
               tibble::as_tibble(),
             color = 'red',
             size = 1,
             alpha = .5
  )
```

```{r}
# verify sp objects
nyc_hf@bbox %>%  as.vector() %>%
  ggmap::get_stamenmap(zoom = 11) %>%
  ggmap() +
  geom_point(aes(x=Longitude, y=Latitude), 
             data = nyc_hf %>% coordinates() %>% 
               tibble::as_tibble(),
             color = 'blue',
             size = 1,
             alpha = .2
  )
```

```{r}
# For Retail food stores
# verify sf objects
nyc_rf_sf %>% st_union() %>% st_bbox() %>% as.vector() %>%
  ggmap::get_stamenmap(zoom = 11) -> baseMap1

ggmap(baseMap1) +
  geom_point(aes(x=X, y=Y), 
             data = nyc_rf_sf %>% st_coordinates() %>% 
               tibble::as_tibble(),
             color = 'red',
             size = 1,
             alpha = .5
  )
```

```{r}
# verify sp objects
nyc_rf@bbox %>%  as.vector() %>%
  ggmap::get_stamenmap(zoom = 11) %>%
  ggmap() +
  geom_point(aes(x=Longitude, y=Latitude), 
             data = nyc_rf %>% coordinates() %>% 
               tibble::as_tibble(),
             color = 'blue',
             size = 1,
             alpha = .2
  )
```

